---
title: "FDA project"
output:
  pdf_document: default
  html_notebook: default
---


```{r}
library(tidyverse)
library(ggplot2)
library(fda)
library(fda.usc)
```

```{r}
df <- read.csv("/Users/gintareklim/Downloads/temperature.csv")
```

```{r}
head(df)
```

```{r}
# converting temperature from kelvin to celcius

convert_kelvin_to_celsius <- function(temp) 
  {
  celsius <- round(temp - 273.15, 2)
  return(celsius)
  }
```

```{r}
colnames(df)
```

```{r}
df$Vancouver <- convert_kelvin_to_celsius(df$Vancouver)
df$Portland <- convert_kelvin_to_celsius(df$Portland)
df$San.Francisco <- convert_kelvin_to_celsius(df$San.Francisco)
df$Seattle <- convert_kelvin_to_celsius(df$Seattle)
df$Los.Angeles <- convert_kelvin_to_celsius(df$Los.Angeles)
df$San.Diego <- convert_kelvin_to_celsius(df$San.Diego)
df$Las.Vegas <- convert_kelvin_to_celsius(df$Las.Vegas)
df$Phoenix <- convert_kelvin_to_celsius(df$Phoenix)
df$Albuquerque <- convert_kelvin_to_celsius(df$Albuquerque)
df$Denver <- convert_kelvin_to_celsius(df$Denver)
df$San.Antonio <- convert_kelvin_to_celsius(df$San.Antonio)
df$Dallas <- convert_kelvin_to_celsius(df$Dallas)
df$Houston <- convert_kelvin_to_celsius(df$Houston)
df$Kansas.City <- convert_kelvin_to_celsius(df$Kansas.City)
df$Minneapolis <- convert_kelvin_to_celsius(df$Minneapolis)
df$Saint.Louis <- convert_kelvin_to_celsius(df$Saint.Louis)
df$Chicago <- convert_kelvin_to_celsius(df$Chicago)
df$Nashville <- convert_kelvin_to_celsius(df$Nashville)
df$Indianapolis <- convert_kelvin_to_celsius(df$Indianapolis)
df$Atlanta <- convert_kelvin_to_celsius(df$Atlanta)
df$Detroit <- convert_kelvin_to_celsius(df$Detroit)
df$Jacksonville <- convert_kelvin_to_celsius(df$Jacksonville)
df$Charlotte <- convert_kelvin_to_celsius(df$Charlotte)
df$Miami <- convert_kelvin_to_celsius(df$Miami)
df$Pittsburgh <- convert_kelvin_to_celsius(df$Pittsburgh)
df$Toronto <- convert_kelvin_to_celsius(df$Toronto)
df$Philadelphia <- convert_kelvin_to_celsius(df$Philadelphia)
df$New.York <- convert_kelvin_to_celsius(df$New.York)
df$Montreal <- convert_kelvin_to_celsius(df$Montreal)
df$Boston <- convert_kelvin_to_celsius(df$Boston)
df$Beersheba <-convert_kelvin_to_celsius(df$Beersheba)
df$Tel.Aviv.District <- convert_kelvin_to_celsius(df$Tel.Aviv.District)
df$Eilat <- convert_kelvin_to_celsius(df$Eilat)
df$Haifa <- convert_kelvin_to_celsius(df$Haifa)
df$Nahariyya <- convert_kelvin_to_celsius(df$Nahariyya)
df$Jerusalem <- convert_kelvin_to_celsius(df$Jerusalem)
```

```{r}
#temperatures converted to from Kelvin to Celcius
head(df)
```

```{r}
#removing first row, since only one city (Eilat) has temperature
df=df[-1,]
```

```{r}
# creating day, month and year columns from column datetime

str(df)
#datetime is a string
```
```{r}
#converted from string to data type
df$date <- as.Date(df$datetime, "%Y-%m-%d %H:%M:%S")
str(df$date)
```

```{r}
df$year <- format(as.Date(df$date, format="%Y-%m-%d"),"%Y")
df$month <- format(as.Date(df$date, format="%Y-%m-%d"),"%m")
df$year_month <- format(as.Date(df$date, format="%Y-%m-%d"),"%Y-%m")
```

```{r}
summary(df)
```

```{r}
dim(df)
```

```{r}
# aggregating to monthly data

monthly1 <- plyr::ddply(df, "year_month", summarise, 
                           Vancouver = round(mean(Vancouver),2))

monthly2 <- plyr::ddply(df, "year_month", summarise, 
                        Portland = round(mean(Portland),2))

monthly3 <- plyr::ddply(df, "year_month", summarise, 
                        San.Francisco = round(mean(San.Francisco),2))

monthly4 <- plyr::ddply(df, "year_month", summarise, 
                        Seattle = round(mean(Seattle),2))

monthly5 <- plyr::ddply(df, "year_month", summarise, 
                        Los.Angeles = round(mean(Los.Angeles),2))

monthly6 <- plyr::ddply(df, "year_month", summarise, 
                        San.Diego = round(mean(San.Diego),2))

monthly7 <- plyr::ddply(df, "year_month", summarise, 
                        Las.Vegas = round(mean(Las.Vegas),2))

monthly8 <- plyr::ddply(df, "year_month", summarise, 
                        Phoenix = round(mean(Phoenix),2))

monthly9 <- plyr::ddply(df, "year_month", summarise, 
                        Albuquerque = round(mean(Albuquerque),2))

monthly10 <- plyr::ddply(df, "year_month", summarise, 
                        Denver = round(mean(Denver),2))

monthly11 <- plyr::ddply(df, "year_month", summarise, 
                       San.Antonio = round(mean(San.Antonio),2))

monthly12 <- plyr::ddply(df, "year_month", summarise, 
                         Dallas = round(mean(Dallas),2))

monthly13 <- plyr::ddply(df, "year_month", summarise, 
                        Houston = round(mean(Houston),2))

monthly14 <- plyr::ddply(df, "year_month", summarise, 
                        Kansas.City = round(mean(Kansas.City),2))

monthly15 <- plyr::ddply(df, "year_month", summarise, 
                        Minneapolis = round(mean(Minneapolis),2))

monthly16 <- plyr::ddply(df, "year_month", summarise, 
                        Saint.Louis = round(mean(Saint.Louis),2))

monthly17 <- plyr::ddply(df, "year_month", summarise, 
                        Chicago = round(mean(Chicago),2))

monthly18 <- plyr::ddply(df, "year_month", summarise, 
                        Nashville = round(mean(Nashville),2))

monthly19 <- plyr::ddply(df, "year_month", summarise, 
                         Indianapolis = round(mean(Indianapolis),2))

monthly20 <- plyr::ddply(df, "year_month", summarise, 
                         Atlanta = round(mean(Atlanta),2))

monthly21 <- plyr::ddply(df, "year_month", summarise, 
                         Saint.Louis = round(mean(Saint.Louis),2))

monthly22 <- plyr::ddply(df, "year_month", summarise, 
                         Detroit = round(mean(Detroit),2))

monthly23 <- plyr::ddply(df, "year_month", summarise, 
                         Jacksonville = round(mean(Jacksonville),2))

monthly24 <- plyr::ddply(df, "year_month", summarise, 
                         Charlotte = round(mean(Charlotte),2))

monthly25 <- plyr::ddply(df, "year_month", summarise, 
                         Miami = round(mean(Miami),2))

monthly26 <- plyr::ddply(df, "year_month", summarise, 
                         Pittsburgh = round(mean(Pittsburgh),2))

monthly27 <- plyr::ddply(df, "year_month", summarise, 
                         Torronto = round(mean(Toronto),2))

monthly28 <- plyr::ddply(df, "year_month", summarise, 
                         Philadelphia = round(mean(Philadelphia),2))

monthly29 <- plyr::ddply(df, "year_month", summarise, 
                         New.York = round(mean(New.York),2))

monthly30 <- plyr::ddply(df, "year_month", summarise, 
                         Montreal = round(mean(Montreal),2))

monthly31 <- plyr::ddply(df, "year_month", summarise, 
                         Boston = round(mean(Boston),2))

monthly30 <- plyr::ddply(df, "year_month", summarise, 
                         Beersheba = round(mean(Beersheba),2))

monthly31 <- plyr::ddply(df, "year_month", summarise, 
                         Tel.Aviv.District = round(mean(Tel.Aviv.District),2))

monthly32 <- plyr::ddply(df, "year_month", summarise, 
                         Eilat = round(mean(Eilat),2))

monthly33 <- plyr::ddply(df, "year_month", summarise, 
                         Haifa = round(mean(Haifa),2))

monthly34 <- plyr::ddply(df, "year_month", summarise, 
                         Nahariyya = round(mean(Nahariyya),2))

monthly35 <- plyr::ddply(df, "year_month", summarise, 
                         Jerusalem = round(mean(Jerusalem),2))
```


```{r}
sum(is.na(df$Vancouver)) #794 missing values
```

```{r}
# merging dataframes

month_avg <- Reduce(function(x, y) merge(x, y, all=TRUE),
       list(monthly1, monthly2, monthly3, monthly4, monthly5,
        monthly6, monthly7, monthly8, monthly9, monthly10,
        monthly11, monthly12, monthly13, monthly14, 
        monthly15, monthly16, monthly17, monthly18,
        monthly19, monthly20, monthly21, monthly22,
        monthly23, monthly24, monthly25, monthly26,
        monthly27, monthly28, monthly29, monthly30, 
        monthly31, monthly32, monthly33, monthly34, 
        monthly35)
       )
```

```{r}
str(month_avg)
```

```{r}
library(zoo)
```
```{r}
#adding year column
month_avg$year <- format(as.Date(as.yearmon(month_avg$year_month)), "%Y")
#ading month column
month_avg$month <- format(as.Date(as.yearmon(month_avg$year_month)), "%m")
```

```{r}
#temperature over years 

ggplot(month_avg, aes(x=month, y=Vancouver, color=as.factor(year))) +
  geom_line(na.rm=TRUE)+geom_point(na.rm=TRUE)+
  labs(title = "Monthly temperature in Vancouver over years", x = "Date", y = "Temperature C",
       color = "Year")
```

```{r}
ggplot(df, aes(x=date, y=Vancouver, color=as.factor(year))) +
  geom_point(na.rm=TRUE)+
  labs(title = "Hourly temperature in Vancouver over years", x = "Date", y = "Temperature C",
       color = "Year")
```

```{r}
ggplot(df, aes(x=date, y=Jerusalem, color=as.factor(year))) +
  geom_point(na.rm=TRUE)+
  labs(title = "Hourly temperature in Jerusalem over years", x = "Date", y = "Temperature C",
       color = "Year")
```

```{r}
ggplot(df, aes(x=date, y=Los.Angeles, color=as.factor(year))) +
  geom_point(na.rm=TRUE)+
  labs(title = "Hourly temperature in Los Angeles over years", x = "Date", y = "Temperature C",
       color = "Year")
```

```{r}
ggplot(df, aes(x=date, y=Toronto, color=as.factor(year))) +
  geom_point(na.rm=TRUE)+
  labs(title = "Hourly temperature in Toronto over years", x = "Date", y = "Temperature C",
       color = "Year")
```

```{r}
ggplot(df, aes(x=date, y=San.Diego, color=as.factor(year))) +
  geom_point(na.rm=TRUE)+
  labs(title = "Hourly temperature in San Diego over years", x = "Date", y = "Temperature C",
       color = "Year")
```

```{r}
#reshaping San Diego temperature

#adding year column
monthly6$year <- format(as.Date(as.yearmon(monthly6$year_month)), "%Y")
#ading month column
monthly6$month <- format(as.Date(as.yearmon(monthly6$year_month)), "%m")

monthly6log <- monthly6 %>% mutate(log10diego = log10(San.Diego))

t_sdiego <- tidyr::spread(monthly6log[,-(1:2)], key = year, value=log10diego)
```

```{r}
# droping so columns for San Diego reshaped data
#to prevent errors since there are nanh values
t_sdiego$"2012" <- NULL
t_sdiego$"2017" <- NULL

```

Creating FDA object

Functional data objects are constructed by specifying a set of basis functions and a set of coefficients defining a linear combination of these basis functions. The B-spline basis is used for non-periodic functions. B-spline basis functions are polynomial segments jointed end-to-end at at argument values called knots, breaks or join points.

```{r}
#creating fda object
monRng <- c(1,12)
bspl <- create.bspline.basis(monRng, norder=12)
Lfdobjelect = int2Lfd(12)
monfdPar <- fdPar(bspl, Lfdobjelect, 0.32)
plot(bspl, lwd=2)
```

Discrete observations on one or more curves and for one more more variables are fit with a set of smooth curves, each defined by an expansion in terms of selected basis functions. The fitting criterion is weighted least squares, and smoothness can be defined in terms of a roughness penalty that is specified in a variety of ways.

```{r}
monMatr <- as.matrix(t_sdiego[,-1])
t_sdiego_fda <- smooth.basis(1:12, monMatr, monfdPar)
plot(t_sdiego_fda)
```
Create a functional data object

```{r}
#create fdata object
t_sdiego_fdata <- fdata(t_sdiego_fda$fd)
plot(t_sdiego_fdata)
```
Functional data estimation via basis representation using generalized cross-validation (GCV) method with a roughness penalty.

```{r}
# optimal smoothing

l <- c(0 , 2^seq(-3, 9, len = 100))
nb <- seq(4, 12, by = 1)
t_sdiego_param <- optim.basis(t_sdiego_fdata, lambda = l, numbasis = nb)
plot(t_sdiego_param$fdata.est)
```

```{r}
t_sdiego_param$gcv.opt
t_sdiego_param$numbasis.opt
t_sdiego_param$lambda.opt
```

Smoothing of functional data using nonparametric kernel estimation with generalized cross-validation (GCV) methods.

```{r}
t_sdiego_np <- optim.np(t_sdiego_fdata, type.S = S.NW, par.CV = list(criteria = "GCV"))
plot(t_sdiego_np$fdata.est)
```

The Fraiman and Muniz (FM) depth. It measures the conditional quantile on all points and then integrated. The depth.FM method sorts the depths along the axis x, ie, performs a ranking of depths.

```{r}
#depth
#non optimally smoothed data

# Fraiman-Muniz Depth
out.FM=depth.FM(t_sdiego_fdata,trim=0.1,draw=TRUE)
```

The depth.RP function implements a depth measure based on random projections.
It builds random projections and calculates the functional depth by the Tukey method combining the information of all projections.

```{r}
out.RP=depth.RP(t_sdiego_fdata,trim=0.1,draw=TRUE)
```
FTS is a fast time series library which creates an fts object by specifying index and data.

```{r}
#FTS object

t_sdiego_fts <- fts(x=1:12, y = monMatr, xname = "Month", yname = "Temperature C")
plot(t_sdiego_fts, plot.type = "functions", plotlegend = TRUE)
```
```{r}
plot(t_sdiego_fts, plot.type="density",plotlegend=TRUE)
```

Compute functional HDR boxplot.

```{r}
fboxplot(t_sdiego_fts, plot.type="bivariate", type="hdr", alpha=c(0.07,0.5),
         projmethod="PCAproj")
```

Visualizing patterns of functional time series. They considered a set of curves as a two-way (p * n) data matrix, where p is the total number of covariates and n is the total number of curves.

The main advantage of this dynamic plot is to visualize both column and row information of a two-way matrix simultaneously, relate the matrix to the corresponding curves, show local variations, and highlight interactions between columns and rows of a two-way matrix.

```{r}
SVDplot(t_sdiego_fts, order = 3, plot = TRUE)
```

Functional outlier detection (abnormality detection) methods.

```{r}
foutliers(t_sdiego_fts, method = "robMah")
```

The outlier detection method corresponds to the approach of Rousseeuw and Leroy (1987) using the robust Mahalanobis distance.

```{r}
foutliers(t_sdiego_fts, method = "lrt")
```

The outlier detection method corresponds to the approach of Febrero et al. (2007) using the likelihood ratio test.

```{r}
foutliers(t_sdiego_fts, method = "depth.trim")
```

The outlier detection method corresponds to the approach of Febrero et al. (2008) using the functional depth with trimmed curves.

```{r}
foutliers(t_sdiego_fts, method = "depth.pond")
```

The outlier detection method corresponds to the approach of Febrero et al. (2008) using the functional depth with all curves.

```{r}
foutliers(t_sdiego_fts, method = "HUoutliers")
```

The outlier detection method corresponds to the approach of Hyndman and Ullah (2008) using the integrated square forecast errors.

FTA to FDATA

```{r}
#FTA to FDATA
t_sdiego_fdata_fts <- fdata(t_sdiego_fts)
class(t_sdiego_fdata_fts)
```

If formats are assigned to the "format" attributes of the data frame columns, the fdata function will apply those formats to the specified columns, and return a new, formatted data frame.

```{r}
l <- c(0 , 2^seq(-3, 9, len = 100))
nb <- seq(4, 12, by = 1)
t_sdiego_param <- optim.basis(t_sdiego_fdata_fts, lambda = l, numbasis = nb)
plot(t_sdiego_param$fdata.est)
```
Optimization of functions using optim.basis was performed applying generalized cross-validation (GCV) method with a roughness penalty.

```{r}
t_sdiego_param$gcv.opt
t_sdiego_param$numbasis.opt
t_sdiego_param$lambda.opt
```

Smoothing of functional data using nonparametric kernel estimation with generalized cross-validation (GCV) methods.

```{r}
t_sdiego_np2 <- optim.np(t_sdiego_fdata_fts, type.S = S.NW, par.CV = list(criteria = "GCV"))
plot(t_sdiego_np2$fdata.est)
```

```{r}
t_sdiego_np2$gcv.opt
```

```{r}
timerange <- seq(1, 12, length=301)
```

Time series

```{r}
#converting fdata class object into fd class object
t_sdiego1_fd <- fdata2fd(t_sdiego_param$fdata.est)

#estimation
t_sdiego1_eval <- predict(t_sdiego1_fd, timerange)

t_sdiego2_fd <- fdata2fd(t_sdiego_np2$fdata.est)

t_sdgiego2_eval <- predict(t_sdiego2_fd, timerange)

t_sdiego1_fts <- fts(x=timerange, y = t_sdiego1_eval, xname = "Month", yname = "Temperature C")

#ploting time series
plot(t_sdiego1_fts, plot.type = "functions", plotlegend = TRUE)
```

```{r}
plot(t_sdiego1_fts, plot.type="density",plotlegend=TRUE)
```
#Principal component score
```{r}
fboxplot(t_sdiego1_fts, plot.type="bivariate", type="bag", projmethod = "PCAproj")
```

```{r}
fboxplot(t_sdiego1_fts, plot.type = "functional", type = "bag", projmethod = "PCAproj")

```

```{r}
fboxplot(t_sdiego1_fts, plot.type="bivariate", type="hdr", alpha=c(0.07,0.5),
         projmethod="PCAproj")
```

Visualizing patterns of functional time series. 

```{r}
SVDplot(t_sdiego1_fts, order = 3, plot = TRUE)
```

```{r}
t_sdiego2_fts <- fts(x=timerange, y = t_sdgiego2_eval, xname = "Month", yname = "Temperature C")
par(mfrow=c(1,1))
plot(t_sdiego2_fts, plot.type = "functions", plotlegend = TRUE)
```
```{r}
plot(t_sdiego2_fts, plot.type="depth",plotlegend=TRUE)
```
```{r}
plot(t_sdiego2_fts, plot.type="density",plotlegend=TRUE)
```

```{r}
fboxplot(t_sdiego2_fts, plot.type="bivariate", type="bag", projmethod = "PCAproj")
```

```{r}
fboxplot(t_sdiego2_fts, plot.type = "functional", type = "bag", projmethod = "PCAproj")
```

```{r}
fboxplot(t_sdiego2_fts, plot.type="bivariate", type="hdr", alpha=c(0.07,0.5),
         projmethod="PCAproj")
```

```{r}
par(mar = rep(2, 4))
SVDplot(t_sdiego2_fts, order = 3, plot = TRUE)
```

```{r}
foutliers(t_sdiego1_fts, method = "robMah")
```

```{r}
foutliers(t_sdiego2_fts, method = "robMah")
```

